When does vapor pressure deficit drive or reduce evapotranspiration?

### 2018-01-08 doc/paper/vpd_et_paper.tex
finished spell check

Answering the question ``When does vapor pressure deficit drive or reduce evapotranspiration?'' is motivated by two potenential but opposing perspectives on the matter. The first persepctive is that increases to vapor pressure deficit (VPD) increases atmospheric demand for water from the land surface, and this drives an increase in evapotranspiration (ET). This perspective is particularly relevant because potential evapotranspiration (PET), which is used in many drought indicies and hydrometeorological studies, only quantifies atmospheric demand effects. However, there is another possible perspective for vegetated surfaces. Plants' stomata have evolved to optimally regulate the exchange of water and carbon, and tend to close in response to increased atmospheric dryness \citep{Ball_1987, Leuning_1990, MEDLYN_2011}.  Therefore, an increase in VPD may actually correspond to a decrease in ET because stomatal closure can cancel the effects of shifts to atmospheric de. So really a question of ``When does VPD drive or reduce ET?'' becomes a question of ``When does plant response or atmospheric demand dominate?.''

Quantifying the plant response to a perturbation to atmospheric VPD increases our understanding of land surface response to shifts in atmospheric conditions. If plant response reduces ET in response to atmospheric drying then soil moisture will be conserved for later ET. An increase in ET in reponse to atmospheric drying will reduce soil moisture, but contribute increased moisenting to the atmosperhere. Understanding the one way response of the land surface to the atmosphere is a critical first step to disentangling land-atmosphere feedbacks. The sign and magnitude of land-surface response drives the co-evolution of the atmosphere and land-surface at many timescales, from diurnal to interdecadal.

@@

\section{Data}
\label{data}
We use data from FLUXNET2015, including all sites with more than four years of data. Each site's plant functional type (PFT) was classified using the International Geosphere-Biosphere Programme vegetation classification scheme (\cite{Loveland_1999}).  Because physical constats used in Section \ref{methods} are only published for five plant functional types (PFTs - see Table \ref{pft}),  only 56 of the 77 sites were used. Figure \ref{map_fig} \explain{map needs to be improved - it's a placeholder for now} presents each site and its plant functional type.

The purpose of this study is to examine ecosystem response to atmospheric drying during the growing season. To accomplish this, we filter and quality control the data using a similiar procedure as \cite{Zhou_2015}:
\begin{itemize}
\item Only measured or highest (``good'') quality gapfilled data, according to qualtiy control flags, are used.
\item To isolate the growing season, we only use days in which the average GPP exceeds 10\% of the observed 95th percentile of GPP for a given site.
\item We remove days with rain and the day following to avoid issues with rain interveption and sensor satruation and high relative humidity (Medlyn et. al. 2017).
\end{itemize}
Additionally, we restrict data to the daytime which is identified when downwelling shortwave radation is greater than 50 W m$^{-2}$ and sensible heat flux is greater than 5 W m$^{-2}$. To further reduce the chance of sensor saturation at high relative humidity, we remove all time steps for which VPD is less than .01 kPa. Timesteps with negative observed GPP or ET are also removed. After these quality control proceedures, 332556 upscaled hourly observations remain.

@@

\begin{table}
  \label{pft}
\caption{Plant functional types, their abbreviation, Medlyn coefficient \citep[from ][]{Lin_2015}, and uWfUE \citep[from ][]{Zhou_2015}. Note that units are converted such that the quantities fit into Equations 1-8 with the variables in Table 1.}
\centering
\begin{tabular}{l c c c}
  \hline
  Abbreviation & PFT & $g_1$ (Pa$^{0.5}$) & uWUE ($\mu$-mol [C] Pa$^{0.5}$ J$^{-1}$ [ET])  \\
  \hline
  CRO & cropland & 183.1 & 3.80 \\
  CSH & closed shrub & 148.6 & 2.18 \\
  DBF & deciduous broadleaf forest & 140.7 & 3.12 \\
  ENF & evergreen needleleaf forest & 74.3 & 3.30 \\
  GRA & grassland (C3) & 166.0 & 2.68 \\
\hline
\multicolumn{2}{l}{$^{a}$Footnote text here.}
\end{tabular}
\end{table}

\begin{figure}[h]
\centering
\includegraphics[width=20pc]{./fig01.pdf}
\caption{Plant functional type and location of sites used in analysis. ***\textbf{This is just  aplaceholder for now, I am currently fixing it***}}
\label{map_fig}
 \end{figure}

@@

Bottom: The location of the minima of ET, as a function of VPD and $\sigma$. Lines and stars denote the distribution of VPD and $\sigma$ next to each axis, following the same percentiles as above. \textbf{***need to replot with correct sigma (normalzied to 1)***}}
\label{term3}
\end{figure}

@@

Figure \ref{real} presents calculated $\frac{\partial \; ET}{\partial \; VPD}$ where, unless otherwise noted, all variables in Equation \ref{d_et} are allowed to vary, including uncertainty. Each column is a different quantity related to $\frac{\partial \; ET}{\partial \; VPD}$, and each row is a different PFT. Vertical black lines denote where our theory says $VPD_{crit}$ should be. For any obervations to the left of this line $\frac{]aprial \; ET}{\partial \; VPD}$ should be negative, and any ovservations the righ right of the line should be postiive. In addition to seeing how the theory $VPD_{crit}$ matches observations, these plots are also useful for observing how the scaling term affects the magnitude of the derivative, if at all.

@@

\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{./agu_fig.pdf}
\caption{Comparison of the sign term with moden uncertainty included (box plots) to the sign term as calculated with simplifying assumptions (blue line).}
\label{agu_fig}
\end{figure}

>>>>

Answering the question ``When does vapor pressure deficit drive or reduce evapotranspiration?'' is motivated by two potential but opposing perspectives on the matter. The first perspective is that increases to vapor pressure deficit (VPD) increases atmospheric demand for water from the land surface, and this drives an increase in evapotranspiration (ET). This perspective is particularly relevant because potential evapotranspiration (PET), which is used in many drought indices and hydrometeorological studies, only quantifies atmospheric demand effects. However, there is another possible perspective for vegetated surfaces. Plants' stomata have evolved to optimally regulate the exchange of water and carbon, and tend to close in response to increased atmospheric dryness \citep{Ball_1987, Leuning_1990, MEDLYN_2011}.  Therefore, an increase in VPD may actually correspond to a decrease in ET because stomatal closure can cancel the effects of shifts to atmospheric demand. So really a question of ``When does VPD drive or reduce ET?'' becomes a question of ``When does plant response or atmospheric demand dominate?.''

Quantifying the plant response to a perturbation to atmospheric VPD increases our understanding of land surface response to shifts in atmospheric conditions. If plant response reduces ET in response to atmospheric drying then soil moisture will be conserved for later ET. An increase in ET in response to atmospheric drying will reduce soil moisture, but contribute increased moistening to the atmosphere. Understanding the one way response of the land surface to the atmosphere is a critical first step to disentangling land-atmosphere feedbacks. The sign and magnitude of land-surface response drives the co-evolution of the atmosphere and land-surface at many timescales, from diurnal to interdecadal.

@@

\section{Data}
\label{data}
We use data from FLUXNET2015, including all sites with more than four years of data. Each site's plant functional type (PFT) was classified using the International Geosphere-Biosphere Programme vegetation classification scheme (\cite{Loveland_1999}).  Because physical constants used in Section \ref{methods} are only published for five plant functional types (PFTs - see Table \ref{pft}),  only 56 of the 77 sites were used. Figure \ref{map_fig} \explain{map needs to be improved - it's a placeholder for now} presents each site and its plant functional type.

The purpose of this study is to examine ecosystem response to atmospheric drying during the growing season. To accomplish this, we filter and quality control the data using a similar procedure as \cite{Zhou_2015}:
\begin{itemize}
\item Only measured or highest (``good'') quality gapfilled data, according to quality control flags, are used.
\item To isolate the growing season, we only use days in which the average GPP exceeds 10\% of the observed 95th percentile of GPP for a given site.
\item We remove days with rain and the day following to avoid issues with rain interception and sensor saturation and high relative humidity (\cite{MEDLYN2016).
\end{itemize}
Additionally, we restrict data to the daytime which is identified when downwelling shortwave radation is greater than 50 W m$^{-2}$ and sensible heat flux is greater than 5 W m$^{-2}$. To further reduce the chance of sensor saturation at high relative humidity, we remove all time steps for which VPD is less than .01 kPa. Timesteps with negative observed GPP or ET are also removed. After these quality control proceedures, 332556 upscaled hourly observations remain.

@@

\begin{table}
  \label{pft}
\caption{Plant functional types, their abbreviation, Medlyn coefficient \citep[from ][]{Lin_2015}, and uWUE \citep[from ][]{Zhou_2015}. Note that units are converted such that the quantities fit into Equations 1-8 with the variables in Table 1.}
\centering
\begin{tabular}{l c c c}
  \hline
  Abbreviation & PFT & $g_1$ (Pa$^{0.5}$) & uWUE ($\mu$-mol [C] Pa$^{0.5}$ J$^{-1}$ [ET])  \\
  \hline
  CRO & cropland & 183.1 & 3.80 \\
  CSH & closed shrub & 148.6 & 2.18 \\
  DBF & deciduous broadleaf forest & 140.7 & 3.12 \\
  ENF & evergreen needleleaf forest & 74.3 & 3.30 \\
  GRA & grassland (C3) & 166.0 & 2.68 \\
\hline
\multicolumn{2}{l}{$^{a}$Footnote text here.}
\end{tabular}
\end{table}

\begin{figure}[h]
\centering
\includegraphics[width=20pc]{./fig01.pdf}
\caption{Plant functional type and location of sites used in analysis. ***\textbf{This is just  a placeholder for now, I am currently fixing it***}}
\label{map_fig}
 \end{figure}

@@

Bottom: The location of the minima of ET, as a function of VPD and $\sigma$. Lines and stars denote the distribution of VPD and $\sigma$ next to each axis, following the same percentiles as above. \textbf{***need to replot with correct sigma (normalized to 1)***}}
\label{term3}
\end{figure}

@@

Figure \ref{real} presents calculated $\frac{\partial \; ET}{\partial \; VPD}$ where, unless otherwise noted, all variables in Equation \ref{d_et} are allowed to vary, including uncertainty. Each column is a different quantity related to $\frac{\partial \; ET}{\partial \; VPD}$, and each row is a different PFT. Vertical black lines denote where our theory says $VPD_{crit}$ should be. For any obervations to the left of this line $\frac{]\partial \; ET}{\partial \; VPD}$ should be negative, and any ovservations the righ right of the line should be postiive. In addition to seeing how the theory $VPD_{crit}$ matches observations, these plots are also useful for observing how the scaling term affects the magnitude of the derivative, if at all.

@@

\begin{figure}[h]
\centering
\includegraphics[width=\textwidth]{./agu_fig.pdf}
\caption{Comparison of the sign term with model uncertainty included (box plots) to the sign term as calculated with simplifying assumptions (blue line).}
\label{agu_fig}
\end{figure}
